---
title: "Comparing calibration of multiple models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing calibration of multiple models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(survival.calib)
```

# Overview



```{r setup, message=FALSE}

library(survival.calib)
library(survival)
library(riskRegression)
library(dplyr)
library(ggplot2)

knitr::opts_chunk$set(fig.width=7, fig.height=5)

```

## Data

For this example, we'll use a standard dataset, `survival::flchain`, with some very light modifications.

```{r}

# drop rows with missing values for simplicity
data_init <- na.omit(flchain)

data_init$chapter <- NULL

head(data_init)

```

We'll also use a simple split of the data for training and testing.

```{r}

n_obs_total <- nrow(data_init)
n_obs_train <- round(n_obs_total * 2/3)

set.seed(32987)
train_index <- sample(n_obs_total, size = n_obs_train)

data_train <- data_init[train_index, ]
data_test <- data_init[-train_index, ]

```

## Model fitting

Suppose we have two model specifications that we want to compare:

- Model 1: use `trt`, `sex`, and `stage` to predict risk for mortality

- Model 2: use `age`, `bili` and `platelet` to predict risk for mortality.

Code to fit these models to the training data is below:

Code to fit these models to the training data is below:

```{r}

model_1 <- coxph(Surv(futime, death) ~ age + sex, 
                 data = data_train,
                 x = TRUE)

model_2 <- update(model_1, . ~ . + lambda + kappa + creatinine)

# compute predicted risk at 3000 days post baseline
time_predict <- 3000

predrisk_1 = predictRisk(model_1, newdata = data_test, times = time_predict)
predrisk_2 = predictRisk(model_2, newdata = data_test, times = time_predict)

calslope_1 <- calib_slope_hare(predrisk_1, 
                               event_time = data_test$futime,
                               event_status = data_test$death,
                               time_predict = time_predict)

calslope_2 <- calib_slope_hare(predrisk_2, 
                               event_time = data_test$futime,
                               event_status = data_test$death,
                               time_predict = time_predict)

calslope <- calib_slope_hare(list(predrisk_1, predrisk_2), 
                             event_time = data_test$futime,
                             event_status = data_test$death,
                             time_predict = time_predict)

caltest <- calib_test_gnd(list(predrisk_1, 
                               predrisk_2, 
                               predrisk_2), 
                          event_time = data_test$futime,
                          event_status = data_test$death,
                          time_predict = time_predict)


```

Now let's combine the data and create a graph showing both curves.

```{r}

data_calslope <- bind_rows(`Model 1` = calslope_hare_1$data,
                           `Model 2` = calslope_hare_2$data,
                           .id = 'model')

fig_cal_slopes <- ggplot(data_calslope) + 
  aes(x = predicted, y = observed, color = model) +
  geom_line() + 
  geom_abline(col = 'grey', linetype = 2, intercept = 0, slope = 1) + 
  scale_x_continuous(limits = c(0,1), 
                     expand = c(0,0),
                     breaks = seq(0, 1, by = 0.2),
                     labels = paste0(seq(0, 100, by = 20),"%")) + 
  scale_y_continuous(limits = c(0, 1), 
                     breaks = seq(0, 1, by = 0.2),
                     labels = paste0(seq(0, 100, by = 20),"%")) +
  theme_bw() + 
  theme(panel.grid = element_blank())

fig_cal_slopes

```

It's standard practice to show the distribution of predicted risk underneath the calibration slope plots. For this visualization, we can use `predrisk_bin_segments`, which creates a dataframe containing all of the aesthetics required for `geom_segment()`. 

```{r}

segment_1 <- predrisk_bin_segments(x = predrisk_1, 
                                   event_status = data_test$status,
                                   event_time = data_test$time,
                                   time_predict = time_predict,
                                   bin_yintercept = -0.05,
                                   bin_length = 1/2)

segment_2 <- predrisk_bin_segments(x = predrisk_2, 
                                   event_status = data_test$status,
                                   event_time = data_test$time,
                                   time_predict = time_predict,
                                   bin_yintercept = -0.15,
                                   bin_length = 1/2)

data_segment <- bind_rows(`Model 1` = segment_1,
                          `Model 2` = segment_2,
                          .id = 'model')

```


```{r, message=FALSE}


fig_cal_slopes + 
  scale_y_continuous(limits = c(-0.2, 1), 
                     breaks = seq(0, 1, by = 0.2),
                     labels = paste0(seq(0, 100, by = 20),"%")) +
  theme(legend.position = c(0.9, 0.33)) +
  geom_hline(yintercept = -0.05, linetype = 2, color = 'grey') +
  geom_hline(yintercept = -0.15, linetype = 2, color = 'grey') +
  geom_segment(data = data_segment, 
               inherit.aes = FALSE,
               size = 1.5,
               mapping = aes(x = x, 
                             y = y,
                             color = model,
                             xend = xend, 
                             yend = yend)) + 
  labs(color = '')

```
